---
title: "609 Final Project"
author: "Christina Taylor"
date: "April 11, 2017"
output: html_document
---

A farmer has 30 acres on which to grow tomatoes and corn. Each 100 bushels of tomatoes require 1000 gallons of water and 5 acres of land. Each 100 bushels of corn require 6000 gallons of water and 2.5 acres of land. Labor costs are $1 per bushel for both corn and tomatoes. The farm has available 30,000 gallons of water and $750 in capital. He knows he cannot sell more than 500 bushels of tomatoes and 475 bushels of corn. He estimates a profit of $2 on each bushel of tomatoes and $3 on each bushel of corn.

(a) How many bushels should he raise to maximize profits?

Let
$t$ = bushels of tomatoes (in hundreds)
$c$ = bushels of corn (in hundreds)
$w$ = gallons of water (in thousands)
$p$ = profit (in dollars)
$a$ = land (in acres)

The farmer's problem can be modeled as 
Maximizing Profit: $200t + 300c$

Subject to the following constraints:
Non-negativity: $t \geq 0, c \geq 0$
Water: $t + 6c \leq 30$
Land: $5t + 2.5c \leq 30$
Labor cost: $t + c \leq 7.5$
Demand: $t \leq 5, c \leq 4.75$

Apply linear programming solution
```{r}
library(lpSolve)
library(lpSolveAPI)

#create LP object with 2 decision variables
lprec = make.lp(0,2)
#set objective function and direction (maximize)
set.objfn(lprec,c(200,300))
lp.control(lprec,sense='max')

#add constraints
add.constraint(lprec, c(1,6), "<=", 30)
add.constraint(lprec, c(5,2.5), "<=", 30)
add.constraint(lprec, c(1,1), "<=", 7.5)
#set boundaries
set.bounds(lprec, lower = c(0,0), upper = c(5,4.75))

#rename for printing
RowNames <- c("water", "land", "labor")
ColNames <- c("tomatoes","corn")
dimnames(lprec) <- list(RowNames, ColNames)
#verify the model
lprec

#solve the model (0 means there is a feasible solution)
solve(lprec)
#find extreme point
get.variables(lprec)
#find maximize objective function value
get.objective(lprec)
```

We can see that if the farmer grows 300 bushels of tomatoes and 450 bushels of corn, he can maximize his profit at $1950. 

(b) Assume the farmer has the oppertunity to sign a contract with grocery store to grow and deliver at least 300 bushels of tomatoes and 500 bushels of corn. Should he sign the contract?

The new demand constraint is now $t \geq 3, c \geq 5$
Intuitively, we can easily see that this condition is incompatible with the constraint of labor costs. We can verify this using linear programming: 
```{r}
lprec = make.lp(0,2)
#set objective function and direction (maximize)
set.objfn(lprec,c(200,300))
lp.control(lprec,sense='max')

#add constraints
add.constraint(lprec, c(1,6), "<=", 30)
add.constraint(lprec, c(5,2.5), "<=", 30)
add.constraint(lprec, c(1,1), "<=", 7.5)
#set boundaries
set.bounds(lprec, lower = c(3,5))

#solve the model
solve(lprec)
```

As shown, there is no feasible solution. The farmer should not sign the contract, because realistically, he cannot deliver. 

(c) Assume the farmer can obtain an additional 10,000 gallons of water for $50. Should he obtain the additional water?

The new water constraint is now $t + 6c \leq 40$
The farmer's profit is $200t + 300c - 50$

```{r}
lprec = make.lp(0,2)
#set objective function and direction (maximize)
set.objfn(lprec,c(200,300))
lp.control(lprec,sense='max')

#add constraints
add.constraint(lprec, c(1,6), "<=", 40)
add.constraint(lprec, c(5,2.5), "<=", 30)
add.constraint(lprec, c(1,1), "<=", 7.5)
#set boundaries
set.bounds(lprec, lower = c(0,0), upper = c(5,4.75))

RowNames <- c("water", "land", "labor")
ColNames <- c("tomatoes","corn")
dimnames(lprec) <- list(RowNames, ColNames)

#solve the model
solve(lprec)
#find extreme point
get.variables(lprec)
#find maximize profit
get.objective(lprec) - 50
```

If the farmer grows 275 bushels of tomatoes and 475 bushels of corn, he can maximize his profit at \$1975. However, given the \$50 cost in acquiring extra water, his net profit actually went down to $1925. Therefore, he should not obtain additional water. 